---
title: "Cheng_Jimmy_Alpha_Diversity"
output: html_document
date: "2026-03-18"
---
## Loading Packages & Phyloseq object
```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggpubr)
library(microbiome)
library(indicspecies)
library(writexl)
library(ggplot2)
```

```{r}
ps = readRDS('phyloseq_taxonomy.rds')

```


## Filter data and group samples
```{r}
psrare = ps %>% rarefy_even_depth(sample.size = 6000, rngseed = 421)


```


```{r}
sd <- data.frame(sample_data(psrare)) 


#group samples into oral and non-oral
sd <- sd %>%            
  mutate(administration = case_when(
    administration %in% c("Infused", "Subcutaneous") ~ "Non-Oral",
    TRUE ~ as.character(administration)
  ))
sd <- sd %>%
  filter(administration %in% c("Oral", "Non-Oral"))  

sd <- filter(sd, !is.na(sd$probiotics))

sd$administration <- as.factor(sd$administration)

sd$probiotics <- as.factor(sd$probiotics)

class(sd$administration)



sample_data(psrare) <- sample_data(sd)


```


## Shannon Alpha Diversity Across Probiotic Status and Administration Routes
```{r}
set.seed(421)

p = plot_richness(psrare,
                  x = "administration",
                  measures = c("Shannon"),
                  color = "administration")
p

pdata = p$data

str(pdata)

p_formatted <- pdata %>%
  ggplot(aes(x = administration, y = value, fill = probiotics)) +
  geom_boxplot(
    aes(group = interaction(administration, probiotics)),
    outlier.shape = NA,
    width = 0.6,
    position = position_dodge(width = 0.7)
  ) +
  geom_jitter(
    aes(group = probiotics),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.7
    ),
    alpha = 0.4,
    size = 2,
    shape = 16,
    color = "black"
  ) +
  scale_fill_manual(values = c(
    "No" = "#F8766D",
    "Yes" = "#00BFC4"
  )) +
  theme_classic(base_size = 18) +
  ylab("Alpha Diversity") +
  xlab("Administration route") +
  labs(fill = "Probiotics")

p_formatted

ggsave(
  "Shannon_Alpha_Diversity.jpeg",
  plot = p_formatted,
  height = 5, width = 8
)
```



## Simpson Alpha Diversity Across Probiotic Status and Administration Routes
```{r}

set.seed(421)

p = plot_richness(psrare,
                  x = "administration",
                  measures = c("Simpson"),
                  color = "administration")
p

pdata = p$data

str(pdata)

p_formatted <- pdata %>%
  ggplot(aes(x = administration, y = value, fill = probiotics)) +
  geom_boxplot(
    aes(group = interaction(administration, probiotics)),
    outlier.shape = NA,
    width = 0.6,
    position = position_dodge(width = 0.7)
  ) +
  geom_jitter(
    aes(group = probiotics),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.7
    ),
    alpha = 0.4,
    size = 2,
    shape = 16,
    color = "black"
  ) +
  scale_fill_manual(values = c(
    "No" = "#F8766D",
    "Yes" = "#00BFC4"
  )) +
  theme_classic(base_size = 18) +
  ylab("Alpha Diversity") +
  xlab("Administration route") +
  labs(fill = "Probiotics")

p_formatted

ggsave(
  "Simpson_Alpha_Diversity.jpeg",
  plot = p_formatted,
  height = 5, width = 8
)

```


## Observed Features Alpha Diversity Across Probiotic Status and Administration Routes
```{r}
set.seed(421)

p = plot_richness(psrare,
                  x = "administration",
                  measures = c("Observed"),
                  color = "administration")
p

pdata = p$data

str(pdata)

p_formatted <- pdata %>%
  ggplot(aes(x = administration, y = value, fill = probiotics)) +
  geom_boxplot(
    aes(group = interaction(administration, probiotics)),
    outlier.shape = NA,
    width = 0.6,
    position = position_dodge(width = 0.7)
  ) +
  geom_jitter(
    aes(group = probiotics),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.7
    ),
    alpha = 0.4,
    size = 2,
    shape = 16,
    color = "black"
  ) +
  scale_fill_manual(values = c(
    "No" = "#F8766D",
    "Yes" = "#00BFC4"
  )) +
  theme_classic(base_size = 18) +
  ylab("Alpha Diversity") +
  xlab("Administration route") +
  labs(fill = "Probiotics")

p_formatted

ggsave(
  "Observed_Alpha_Diversity.jpeg",
  plot = p_formatted,
  height = 5, width = 8
)


```


## Chao1 Alpha Diversity Across Probiotic Status and Administration Routes

```{r}
set.seed(421)

p = plot_richness(psrare,
                  x = "administration",
                  measures = c("Chao1"),
                  color = "administration")
p

pdata = p$data

str(pdata)

p_formatted <- pdata %>%
  ggplot(aes(x = administration, y = value, fill = probiotics)) +
  geom_boxplot(
    aes(group = interaction(administration, probiotics)),
    outlier.shape = NA,
    width = 0.6,
    position = position_dodge(width = 0.7)
  ) +
  geom_jitter(
    aes(group = probiotics),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.7
    ),
    alpha = 0.4,
    size = 2,
    shape = 16,
    color = "black"
  ) +
  scale_fill_manual(values = c(
    "No" = "#F8766D",
    "Yes" = "#00BFC4"
  )) +
  theme_classic(base_size = 18) +
  ylab("Alpha Diversity") +
  xlab("Administration route") +
  labs(fill = "Probiotics")

p_formatted

ggsave(
  "Chao1_Alpha_Diversity.jpeg",
  plot = p_formatted,
  height = 5, width = 8
)



```



## Four metrics Alpha Diversity Across Probiotic Status and Administration Routes
```{r}

set.seed(421)

p = plot_richness(psrare,
                  x = 'administration',
                  measures = c('Observed','Shannon','Chao1','Simpson'),
                  color = 'administration')
p

pdata = p$data

str(pdata)

p_formatted <- pdata %>%
  ggplot(aes(x = administration, y = value, fill = probiotics)) +
  geom_boxplot(
    aes(group = interaction(administration, probiotics)),
    outlier.shape = NA,
    width = 0.6,
    position = position_dodge(width = 0.7)
  ) +
  geom_jitter(
    aes(group = probiotics),
    position = position_jitterdodge(
      jitter.width = 0.08,
      dodge.width = 0.7
    ),
    alpha = 0.4,
    size = 2,
    shape = 16,
    color = "black"
  ) +
  scale_fill_manual(values = c(
    "No" = "#F8766D",
    "Yes" = "#00BFC4"
  )) +
  theme_classic(base_size = 18) +
  facet_wrap(~variable,ncol=4,scales = 'free_y') +
  ylab("Alpha Diversity") +
  xlab("Administration route") +
  theme(axis.text.x = element_text(angle = 45,vjust=1,hjust=1)) +
  labs(fill = "Probiotics")

p_formatted

ggsave(
  "Shannon_Alpha_Diversity.jpeg",
  plot = p_formatted,
  height = 5, width = 16
)


```



## Linear Regreassion Model of Alpha Diversity with Interaction Effects
```{R}

#run a linear regression model instead of krustcal-Wallis test
stats <- pdata %>%
  group_by(variable) %>%
  group_modify(~{
    temp_stats = data.frame()
    lmmod <- glm(value ~ probiotics * administration, data = .x)
    
    temp_stats <- broom::tidy(lmmod) %>%
      mutate(effect = "interaction") %>%   mutate(padj = p.adjust(p.value, method='BH'), .after='p.value')

    return(temp_stats)
  }) %>%
  ungroup() %>%
  arrange(term)

stats2 <- stats %>%
  filter(term =="probioticsYes:administrationOral")
View(stats2)


View(stats)


maxvals <- pdata %>%
  group_by(variable) %>%
  summarize(max.y = max(value)) %>% ungroup()

annot <- stats2 %>%
  mutate(label = paste('Padj-INT=',signif(padj,digits = 2),sep='')) %>%
  left_join(maxvals, by = "variable") %>%
  mutate(ypos = 1.1 * max.y, xpos = 1.5, probiotics = "Yes", administration = "Oral", "Non-Oral", )

```

##Add statistical annotations to plot
```{r}
p_final <- p_formatted +
  geom_text(data = annot,
            aes(x = xpos, y = ypos, label = label),
            size = 5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

p_final


ggsave('Final plot- Alpha Diversity-simpson.jpeg',
       plot = p_final,
       height= 5, width = 16)


```
