---
title: "beta diversity"
output: html_document
date: "2026-03-27"
---

## Loading Packages & Phyloseq object

```{r}
library(tidyverse)
library(phyloseq)
library(vegan)
library(writexl)
```

```{r}
ps <- readRDS('phyloseq_taxonomy.rds')
```

## Rarefy data and group samples

```{r}
psrare <- ps %>% rarefy_even_depth(sample.size = 6000, rngseed = 421, verbose = FALSE)
```

```{r}

# Group the samples into oral and non-oral, and then divide into four groups based on probiotic use

sd <- data.frame(sample_data(psrare))

sd <- sd %>%
  mutate(administration = case_when(
    administration %in% c("Infused", "Subcutaneous") ~ "Non-Oral",
    TRUE ~ as.character(administration)
  ))

sd <- sd %>%
  filter(administration %in% c("Oral", "Non-Oral"))

sd <- filter(sd, !is.na(sd$probiotics))

sd$administration <- as.factor(sd$administration)
sd$probiotics <- as.factor(sd$probiotics)

class(sd$administration)

#create combined group labeling for samples 
sd <- sd %>%
  mutate(Groups = paste(administration, probiotics, sep = "_"))

sd$Groups <- factor(sd$Groups,
                   levels = c("Non-Oral_No", "Non-Oral_Yes", "Oral_No", "Oral_Yes"))

# Update phyloseq object
sample_data(psrare) <- sample_data(sd)

metadata = sample_data(psrare) %>% data.frame()
```

## Beta diversitry distance matrix

```{r}
# Bray-Curtis
dist_bray <- phyloseq::distance(psrare, method = "bray")
# Jaccard
dist_jaccard <- phyloseq::distance(psrare, method = "jaccard")
# Unweighted UniFrac
dist_unweighted <- phyloseq::distance(psrare, method = "unifrac", weighted = FALSE)
# Weighted UniFrac
dist_weighted <- phyloseq::distance(psrare, method = "unifrac", weighted = TRUE)
```


## Bray-Curtis MDS plot
```{r}
# Bray-Curtis
mds = metaMDS(dist_bray)

mds_data_1 = mds$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data_1)

dist_bray = mds_data_1 %>%
  ggplot(aes(MDS1, MDS2, color = Groups)) +
  geom_point() +
  stat_ellipse(linewidth = 1) +
  scale_color_manual(values = c(
    "Non-Oral_No"  = "#4C78A8",  # original blue
    "Non-Oral_Yes" = "#6BAED6",  # stronger medium blue 
    "Oral_No"      = "#D55E00",  # stronger orange 
    "Oral_Yes"     = "#E69F00"   # orange-yellow
  )) +
  theme_classic(base_size = 14)

dist_bray


ggsave(
  "Bray-Curtis_Beta_Diversity.jpeg",
  plot = dist_bray,
  height = 5, width = 8
)

```

## Jaccard MDS plot
```{r}

# Jaccard

mds_2 = metaMDS(dist_jaccard)

mds_data_2 = mds_2$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data_2)

dist_jaccard = mds_data_2 %>%
  ggplot(aes(MDS1,MDS2,color = Groups)) +
  geom_point() +
  stat_ellipse() +
   scale_color_manual(values = c(
    "Non-Oral_No"  = "#4C78A8",  # original blue
    "Non-Oral_Yes" = "#6BAED6",  # stronger medium blue
    "Oral_No"      = "#D55E00",  # stronger orange
    "Oral_Yes"     = "#E69F00"   # orange-yellow
  )) +
  theme_classic(base_size=14)
dist_jaccard


ggsave(
  "Jaccard_Beta_Diversity.jpeg",
  plot = dist_jaccard,
  height = 5, width = 8
)
```

## Unweighted UniFrac MDS plot
```{r}
# Unweighted UniFrac

mds_3 = metaMDS(dist_unweighted)

mds_data_3 = mds_3$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data_3)

dist_unweighted = mds_data_3 %>%
  ggplot(aes(MDS1,MDS2,color = Groups)) +
  geom_point() +
  stat_ellipse() +
   scale_color_manual(values = c(
    "Non-Oral_No"  = "#4C78A8",  # original blue
    "Non-Oral_Yes" = "#6BAED6",  # stronger medium blue
    "Oral_No"      = "#D55E00",  # stronger orange
    "Oral_Yes"     = "#E69F00"   # orange-yellow
  )) +
  theme_classic(base_size=14)
dist_unweighted


ggsave(
  "Unweighted UniFrac_Beta_Diversity.jpeg",
  plot = dist_unweighted,
  height = 5, width = 8
)
```

## Weighted UniFrac matrix and MDS plot
```{r}
# Weighted UniFrac

mds_4 = metaMDS(dist_weighted)

mds_data_4 = mds_4$points %>% as.data.frame %>%
  merge(sample_data(psrare), by='row.names', sort=F)
head(mds_data_4)

dist_weighted = mds_data_4 %>%
  ggplot(aes(MDS1,MDS2,color = Groups)) +
  geom_point() +
  stat_ellipse() +
   scale_color_manual(values = c(
    "Non-Oral_No"  = "#4C78A8",  # original blue
    "Non-Oral_Yes" = "#6BAED6",  # stronger medium blue
    "Oral_No"      = "#D55E00",  # stronger orange
    "Oral_Yes"     = "#E69F00"   # orange-yellow
  )) +
  theme_classic(base_size=14)
dist_weighted


ggsave('Weighted UniFrac_Beta_Diversity.jpeg',
       plot = dist_bray,
       height= 5, width = 8)



```

## Bray-Curtis PERMANOVA statistical test with interaction term
```{r}

# PERMANOVA with interaction term
stats_bray <- adonis2(
  dist_bray ~ administration * probiotics,
  data = metadata,
  by = "margin"
)

stats_bray


stats = bind_rows('Multivariate' = stats_bray %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 


stats_b = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

stats_b


```

## Jaccard PERMANOVA statistical test with interaction term
```{r}

# PERMANOVA with interaction term - jaccard

stats_jac <- adonis2(
  dist_jaccard ~ administration * probiotics,
  data = metadata,
  by = "margin"
)

stats_jac


stats = bind_rows('Multivariate' = stats_jac %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 



stats_j = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

stats_j 



```

## Unweighted UniFrac PERMANOVA statistical test with interaction term
```{r}
# PERMANOVA with interaction term - unweighted

stats_unweighted <- adonis2(
  dist_unweighted ~ administration * probiotics,
  data = metadata,
  by = "margin"
)

stats_unweighted


stats = bind_rows('Multivariate' = stats_unweighted %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 



stats_un = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval)) 

stats_un 


```

## Weighted UniFrac PERMANOVA statistical test with interaction term
```{r}
# PERMANOVA with interaction term - weighted

stats_weighted <- adonis2(
  dist_weighted ~ administration * probiotics,
  data = metadata,
  by = "margin"
)

stats_weighted


stats = bind_rows('Multivariate' = stats_weighted %>% as.data.frame() %>% 
                    rownames_to_column('Variable'),
                  .id = 'Model') 



stats_weight = stats %>% rename(Pval = `Pr(>F)`) %>%
  filter(!is.na(Pval))

stats_weight 

```
